---
title: "Proposal Project"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## LOAD REQUIDED PACKAGES
```{r Load-packages}

# Load required packages
library(RAQSAPI)
library(purrr)
library(readr)
library(dplyr)
library(geosphere)
library(ggplot2)
library(gridExtra)
library(lubridate)


```


## EPA CREDENTIALS
```{r set_credentials}
# EPA credentials
aqs_credentials(username = "rnl34@case.edu", key = "copperhawk19")
```


## EPA SITE NUMBERS AND DATE RANGE
```{r }

# Vector of site numbers
site_numbers <- c("0034", "0038", "0045", "0060", "0065", "0073")

# Define date range
start_date <- as.Date("2024-11-01")
end_date   <- as.Date("2025-05-01")
```

## DOWNLOAD AND SAVE EPA DATA.

```{r Download-Save EPA Data}
# Loop through and download data for each site
epa_data_list <- map(site_numbers, function(site) {
  aqs_sampledata_by_site(
    parameter = "88101",              # PM2.5
    bdate = start_date,
    edate = end_date,
    stateFIPS = "39",                 # Ohio
    countycode = "035",               # Cuyahoga
    sitenum = site,
    return_header = FALSE
  )
})

# Assign a name to each element that is its site number
names(epa_data_list) <- site_numbers
```

## VIEW AND SAVE SAMPLE DATA

```{r View-and- Save}
# View one dataset (site "0060")
head(epa_data_list[["0060"]])

# Save each to its own CSV file
walk2(epa_data_list, names(epa_data_list), ~ write_csv(.x, paste0("epa_pm25_site_", .y, ".csv")))
```

## COMBINE ALL EPA SITE DATA AND ASSESS MISSING VALUES
```{r}

# Combine all EPA site data into one tibble
epa_all_data <- bind_rows(epa_data_list, .id = "site_number")

# Count NA values in each column
na_counts <- sapply(epa_all_data, function(col) sum(is.na(col)))
na_counts <- sort(na_counts, decreasing = TRUE)

# Display NA counts per column
na_counts

# Identify columns with mostly missing values
total_rows <- nrow(epa_all_data)

# Columns where fewer than 5% of rows are NOT NA
mostly_na_cols <- names(na_counts[na_counts > 0 & na_counts < total_rows * 0.05])

# Summary of columns with mostly missing values
epa_all_data %>%
  select(all_of(mostly_na_cols)) %>%
  summary()

```


```{r prepare_epa_1hr_data}

# list of data frames for each site downloaded from AQS API


# Combine all site data into one data frame
epa_all_sites_raw <- bind_rows(epa_data_list, .id = "site_number")

# Keep only 1-hour PM2.5 measurements with valid values
epa_1hr_clean <- epa_all_sites_raw %>%
  filter(sample_duration == "1 HOUR", !is.na(sample_measurement)) %>%
  mutate(
    datetime_local = as.POSIXct(paste(date_local, time_local),
                                format = "%Y-%m-%d %H:%M",
                                tz = "America/New_York"),
    date = as.Date(datetime_local)
  )


# View structure of the cleaned dataset
glimpse(epa_1hr_clean)

# Count and sort NA values in each column
na_counts_1hr <- sapply(epa_1hr_clean, function(col) sum(is.na(col)))
na_counts_1hr <- sort(na_counts_1hr, decreasing = TRUE)

# Display NA counts
na_counts_1hr
```

```{r}
# Daily average PM2.5 for each site
epa_1hr_daily_avg <- epa_1hr_clean %>%
  group_by(site_number, date) %>%
  summarise(daily_avg_pm25 = mean(sample_measurement, na.rm = TRUE), .groups = "drop")


# Create and save time-series plots for each site
unique_sites <- unique(epa_1hr_daily_avg$site_number)

for (site in unique_sites) {
  site_data <- epa_1hr_daily_avg %>% filter(site_number == site)

  p <- ggplot(site_data, aes(x = date, y = daily_avg_pm25)) +
    geom_line(color = "steelblue", linewidth = 0.6) +
    labs(
      title = paste("Daily Average PM2.5 (1-Hour Data) - Site", site),
      x = "Date",
      y = expression("PM"[2.5]*" (µg/m³)")
    ) +
    scale_x_date(date_breaks = "1 month", date_labels = "%b\n%Y") +
    theme_minimal(base_size = 13)

  print(p)  # Display plot in Viewer pane

  # Save plot as PNG
  ggsave(filename = paste0("pm25_daily_avg_site_", site, ".png"),
         plot = p, width = 8, height = 4.5)
}

```
The daily average PM2.5 concentrations based on 1-hour measurements from six selected EPA monitoring sites in Cuyahoga County between November 2024 and May 2025. Key observations include:

	•	Temporal Pattern: Most sites show noticeable PM2.5 spikes between December 2024 and February 2025, potentially indicating higher pollution during winter months—likely due to increased heating and stagnant atmospheric conditions.
	•	Site 0034 and 0038 generally show moderate fluctuations, with occasional peaks above 15 µg/m³.
	•	Sites 0045, 0065, and 0073 exhibit more pronounced spikes, with some values exceeding 20 µg/m³.
	•	Spring Transition: PM2.5 levels tend to decline across all sites from March to May 2025, suggesting improved air quality as temperatures rise and atmospheric dispersion improves.
	

## 24-hour average PM2.5 data from selected EPA monitoring sites, 
```{r}

# Vector of site numbers
site_numbers <- c("0034", "0038", "0045", "0060", "0065", "0073")

# Read and combine all 24-hour data files
epa_24hr_clean <- bind_rows(
  lapply(site_numbers, function(site) {
    read_csv(paste0("epa_pm25_site_", site, ".csv")) %>%
      filter(sample_duration == "24 HOUR") %>%
      mutate(
        site_number = site,
        date = as.Date(date_local),
        pm25 = sample_measurement
      ) %>%
      select(site_number, date, pm25)
  })
)

library(ggplot2)

unique_sites_24hr <- unique(epa_24hr_clean$site_number)

for (site in unique_sites_24hr) {
  site_data <- epa_24hr_clean %>% filter(site_number == site)

  p <- ggplot(site_data, aes(x = date, y = pm25)) +
    geom_line(color = "firebrick", linewidth = 0.6) +
    labs(
      title = paste("Daily PM2.5 (24-Hour Sample) - Site", site),
      x = "Date",
      y = expression("PM"[2.5]*" (µg/m³)")
    ) +
    scale_x_date(date_breaks = "1 month", date_labels = "%b\n%Y") +
    theme_minimal(base_size = 13)

  print(p)
  ggsave(filename = paste0("pm25_24hr_site_", site, ".png"), plot = p, width = 8, height = 4.5)
}
```
	•	All sites exhibit fluctuations in PM2.5 with noticeable peaks in late December through February, suggesting potential seasonal sources or weather effects.
	•	Sites like 0038 and 0065 show more extreme spikes compared to 0034 or 0073, indicating either more variability in emissions or sensor noise.

## Comparing Daily Averages of 1-Hour PM2.5 with 24-Hour Measurements
```{r}
# Loop over each site
for (site in site_numbers) {
  
  # Load data
  data <- read_csv(paste0("epa_pm25_site_", site, ".csv"))
  
  # Prepare 1-hour data: group and average by date
  hourly_avg <- data %>%
    filter(sample_duration == "1 HOUR", !is.na(sample_measurement)) %>%
    mutate(date = as.Date(date_local)) %>%
    group_by(date) %>%
    summarise(pm25_1hr_avg = mean(sample_measurement, na.rm = TRUE), .groups = "drop")
  
  # Prepare 24-hour data
  daily_data <- data %>%
    filter(sample_duration == "24 HOUR", !is.na(sample_measurement)) %>%
    mutate(date = as.Date(date_local)) %>%
    select(date, pm25_24hr = sample_measurement)
  
  # Join both datasets by date
  comparison_df <- left_join(hourly_avg, daily_data, by = "date") %>%
    filter(!is.na(pm25_1hr_avg), !is.na(pm25_24hr))  # Keep only rows with both values
  
  # Plot X-Y scatter
  p <- ggplot(comparison_df, aes(x = pm25_24hr, y = pm25_1hr_avg)) +
    geom_point(color = "steelblue", alpha = 0.7) +
    geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "darkred") +
    labs(
      title = paste("EPA: 1-Hour Avg vs 24-Hour PM2.5 (Site", site, ")"),
      x = expression("24-Hour PM"[2.5]*" (µg/m³)"),
      y = expression("Daily Avg of 1-Hour PM"[2.5]*" (µg/m³)")
    ) +
    theme_minimal(base_size = 13)
  
  print(p)
}

```

Takeaways from the Plots:
	•	Each plot shows a strong positive linear relationship, confirming that daily averages of 1-hour data align well with 24-hour official readings.
	•	Sites like 0038 and 0060 show particularly tight clustering along the line of best fit, indicating good agreement and minimal deviation.
	•	In contrast, sites like 0045 and 0065 exhibit more scatter, suggesting possible variability in sensor behavior or local conditions affecting hourly data.
	
These comparisons are helpful for:
	•	Validating the consistency of hourly data for estimating daily exposure,
	•	Identifying sites where hourly measurements may be less reliable,
	•	Supporting future sensor calibration efforts.


# Loading CLEANinCLE Data for PM2.5
```{r}

# Load CLEANinCLE raw data
cleanincle_data <- read_csv("~/Documents/R_Projects/Air_pollution_Projects/CLEANinCLE_Raw_Air_Quality_Data_2024-11-2025-05.csv")

# Convert time column and filter for PM2.5 values
cleanincle_data <- cleanincle_data %>%
  filter(measure_name == "mc_pm2_5", !is.na(value)) %>%
  mutate(
    datetime = as.POSIXct(time, format = "%Y-%m-%d %H:%M:%S", tz = "America/New_York"),
    date = as.Date(datetime)
  )

# Preview the cleaned data
glimpse(cleanincle_data)
```

# Daily average PM2.5 from the optical sensor data.
```{r}
# Daily average PM2.5 values from CLEANinCLE
cleanincle_daily_avg <- cleanincle_data %>%
  group_by(date) %>%
  summarise(pm25_avg = mean(value, na.rm = TRUE)) %>%
  ungroup()
```


# Merging CLEANinCLE and EPA data by date.
```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(lubridate)

# Vector of EPA site numbers
site_numbers <- c("0034", "0038", "0045", "0060", "0065","0073")

# Step 1: Calculate CLEANinCLE daily average
cleanincle_daily_avg <- cleanincle_data %>%
  group_by(date) %>%
  summarise(pm25_cleanincle = mean(value, na.rm = TRUE), .groups = "drop")

# Step 2: Loop through each site and create comparison plot
for (site in site_numbers) {
  # Filter EPA data for this site
  epa_24hr_site <- epa_24hr_clean %>%
    filter(site_number == site) %>%
    select(date, pm25_epa = pm25)

  # Join CLEANinCLE with EPA data on date
  comparison_df <- inner_join(cleanincle_daily_avg, epa_24hr_site, by = "date")

  # Plot comparison
  p <- ggplot(comparison_df, aes(x = pm25_cleanincle, y = pm25_epa)) +
    geom_point(color = "steelblue", size = 2, alpha = 0.75) +
    geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 0.8) +
    labs(
      title = paste("Daily PM2.5 Comparison: CLEANinCLE vs EPA 24-Hour (Site", site, ")"),
      x = "CLEANinCLE Daily Avg PM2.5 (µg/m³)",
      y = "EPA 24-Hour PM2.5 (µg/m³)"
    ) +
    theme_minimal(base_size = 13)

  geom_abline(intercept = 0, slope = 1, linetype = "dotted", color = "gray50")
  print(p)

  # Optionally save plot
  ggsave(
    filename = paste0("Daily PM2.5 Comparison: CLEANinCLE vs EPA 24-Hour", site, ".png"),
    plot = p,
    width = 8, height = 5
  )
}

```
Takeaways 
	1.	Positive Correlation Across All Sites
All sites show a positive relationship between CLEANinCLE and EPA PM2.5 values, suggesting that the low-cost sensors generally track trends in pollution correctly.

	2.	Slight Underestimation at Higher Values
In most plots (especially Sites 0038, 0045, and 0060), you can see a clustering of points below the 1:1 trend line at higher PM2.5 concentrations. This suggests the CLEANinCLE sensors may be underestimating PM2.5 levels when pollution is higher. A common issue with some low-cost optical sensors due to humidity or calibration limitations.

	3.	Spread and Consistency Vary by Site
	•	Sites 0034 and 0038: Show stronger and tighter clustering around the trend line. These sensors may be better calibrated or exposed to more consistent pollution sources.
	•	Site 0045: Shows more scatter, which might reflect either noisier sensor readings or more localized pollution events not captured as well by the EPA sensor.
	•	Site 0065: Also shows underprediction at higher levels and has more vertical spread. This could suggest calibration drift or micro-environmental variation.
	•	Site 0073: The spread appears wider and the trend weaker, possibly due to fewer overlapping days or inconsistent readings.
	
	4.	Slope of Regression Line Is < 1 in Most Cases
The red trend line slopes less steeply than a 1:1 line would, again indicating that as true PM2.5 rises, the CLEANinCLE values rise too, but not as much likely due to underestimation bias.


```{r}
epa_sites <- aqs_sites_by_county(
  stateFIPS = "39",    # Ohio
  countycode = "035",  # Cuyahoga County
  return_header = FALSE
)

# Preview to see latitude and longitude
head(epa_sites)

cleanincle_coords <- cleanincle_data %>%
  select(device_eui, latitude, longitude) %>%
  distinct()

# Preview
head(cleanincle_coords)
```


```{r epa-6x6-filter-grid, fig.width=22, fig.height=22, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(gridExtra)

epa_wide <- epa_24hr_clean %>%
  select(date, site_number, pm25) %>%
  pivot_wider(names_from = site_number, values_from = pm25, values_fn = mean) %>%
  filter(rowSums(is.na(.)) < 3)

site_names <- sort(setdiff(names(epa_wide), "date"))
n_sites <- length(site_names)
axis_min <- 0
axis_max <- 28

plot_matrix <- vector("list", n_sites * n_sites)

for (i in seq_len(n_sites)) {
  for (j in seq_len(n_sites)) {
    x_site <- site_names[j]
    y_site <- site_names[i]
    index <- (i - 1) * n_sites + j

    df_pair <- epa_wide %>%
      select(x = all_of(x_site), y = all_of(y_site)) %>%
      filter(!is.na(x) & !is.na(y))

    r2_text <- if (nrow(df_pair) > 2) {
      r2 <- summary(lm(y ~ x, data = df_pair))$r.squared
      paste0("R² = ", round(r2, 2))
    } else {
      "R² = NA"
    }

    p <- ggplot(df_pair, aes(x = x, y = y)) +
      geom_point(color = "steelblue", alpha = 0.7, size = 1.8) +
      geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
      coord_fixed(xlim = c(axis_min, axis_max), ylim = c(axis_min, axis_max)) +
      labs(
        x = paste("Site", x_site),
        y = paste("Site", y_site),
        subtitle = r2_text
      ) +
      theme_minimal(base_size = 12) +
      theme(
        plot.subtitle = element_text(size = 10, hjust = 0.5),
        axis.title = element_text(size = 10),
        axis.text = element_text(size = 9),
        plot.margin = margin(5, 5, 5, 5)
      )

    plot_matrix[[index]] <- p
  }
}

grid_plot <- grid.arrange(grobs = plot_matrix, ncol = n_sites, nrow = n_sites)
print(grid_plot)

ggsave("EPA_Filter_PM25_6x6_Comparison.pdf", plot = grid_plot, width = 22, height = 22)
```
Key Takeaway:

The scatterplot matrix shows strong consistency in daily PM2.5 measurements across many EPA monitoring sites in Cuyahoga County, with most site pairs exhibiting high R² values and tight clustering around the 1:1 line. 
This suggests:
	•	Good agreement among EPA filter-based sites, supporting spatial reliability of the EPA network.
	•	Some variability in correlation strength, possibly due to differences in local emission sources, wind patterns, or sensor performance at specific sites.
	•	Diagonal plots (Site vs. same Site) expectedly show perfect agreement (R² = 1), while a few off-diagonal comparisons show slightly weaker alignment, indicating potential spatial heterogeneity or missing data impacts.
	
```{r}

```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
